---
title: "Juvenile snowshoe hares in Bonanza Creek: exploratory report"
author: "By: Larissa Neilson"
date: "11/16/2020"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(kableExtra)
library(effsize)
library(broom)
library(ggpubr)
```

## Introduction

This report provides an exploratory overview of differences in juvenile snowshoe hare sizes observed at the Bonanza Creek Experimental Forest (Long Term Ecological Research site). Specifically, juvenile weight (g) is compared across sex and site, and the relationship between hind foot length and weight is explored (Kielland et  al. 2017).

## Data and analyses

Bonanza hares structural size measurements were collected at the Bonanza Creek Long Term Ecological Research (LTER) network site. The data contain observations for 3,197 snowshoe hares collected at three sites (Bonanza riparian, Bonanza mature, and a mature lowland Black Spruce Stand) from 1998-2012. Following exploratory data visualization, male and female juvenile snowshoe hare weights were compared by two-sample t-tests using a significance level (α) of 0.05 throughout. Differences between groups are described by Cohen’s d effect size. The relationship between hind foot length and weight is explored by simple linear regression. All analyses are in R version 4.0.2 using RStudio version 1.3.1073.

```{r, include = FALSE}
bonanza_hares <- read.csv(here("data", "bonanza_hares.csv")) %>% 
  clean_names()
```

## Exploratory findings

### Annual juvenile hare trap counts

```{r, include = FALSE}
juvenile_data <- bonanza_hares %>% 
  filter(age == "j") %>% 
  mutate(date_new = mdy(date)) %>%
  mutate(year_new = year(date_new)) %>% 
  mutate(grid_full = case_when(
    grid == "bonbs" ~ "Black Spruce stand",
    grid == "bonmat" ~ "Bonanza mature",
    grid == "bonrip" ~ "Bonanza riparian"
  ))
```

```{r, include = FALSE}
annual_juvenile_counts <- juvenile_data %>% 
  count(year_new)
```

```{r, echo = FALSE, warning = FALSE}
ggplot(data = annual_juvenile_counts, aes(x = year_new, y = n)) +
  geom_col(fill = "darkgreen") +
  labs(x = NULL, y = "Juvenile hare count") +
  scale_x_continuous(breaks = c(1998:2012)) +
  theme_minimal()
```

```{r, include = FALSE, warning = FALSE}
juvenile_count_stats <- annual_juvenile_counts %>%
  summarize(mean_n = mean(n),
            median_n = median(n))
```

**Figure 1.** Annual juvenile snowshoe hare trap counts from 1998-2012. Data: Kielland et  al. (2017).

The maximum juvenile hare trappings occurred in 1999 (n = 126). The minimum trappings occurred in 2010 (n = 2). During the years 1998, 2002, and 2009, there were no juveniles hares trapped at the site. The average annual number of juvenile hares trapped is 31, while the median number of juveniles trapped is 18. This shows that the larger sample sizes seen during some years are positively influencing the mean number of juvenile trappings. Overall, the number of juvenile trappings seems to be declining with time.
Due to the unequal number of days and traps studied each year, these annual trap counts aren't as accurate as they could be. For future analyses, I would recommend standardizing the methodology across the sites to ensure the same number of days and traps are studied each year.


```{r, include = FALSE, warning = FALSE}
# Summary statistics
juvenile_weight_stats <- juvenile_data %>% 
  group_by(sex, grid_full) %>% 
  summarize(mean_weight = mean(weight, na.rm = T))

kbl(juvenile_weight_stats) %>% 
  kable_styling()
```

### Comparison of juvenile hare weights by sex and site

```{r, echo = FALSE, warning = FALSE}
ggplot(data = juvenile_data, aes(x = sex,  y = weight)) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(x = NULL, y = "Weight (grams)") +
  scale_x_discrete(labels = c("Female", "Male", "Unspecified")) +
  facet_wrap(~ grid_full) +
  theme(legend.position = "none") +
  geom_point(data = juvenile_weight_stats, aes(x = sex, y = mean_weight), color = "red")

```

**Figure 2.** Comparison of juvenile snowshoe hare weights by sex and site. Data: Kielland et  al. (2017).

Across the three sites, the juvenile male hares tended to be bigger in terms of weight (grams). The black dots on the jitter plots indicate individual observations while the red dots indicate the average weight of that group. The biggest juvenile hares were found in the Black Spruce stand site with higher average weights of both male and female hares. Additionally, the Bonanza riparian site seemed to have the highest number of observations across the sites; I would recommend standardizing the methodology to have an equal number of observations across the three sites. 

### Juvenile weight comparison by sex

```{r, include = FALSE, warning = FALSE}
# Exploratory graphs
# Histogram
ggplot(data = juvenile_data, aes(x = weight)) +
  geom_histogram(bins = 15) +
  facet_wrap(~ sex)

# QQ-plot
ggplot(data = juvenile_data, aes(sample = weight)) +
  geom_qq()
  facet_wrap(~ sex)
```

**Table 1.** Descriptive statistics (mean, standard deviation, and sample size) for male and female juvenile snowshoe hares. Data: Kielland et  al. (2017).
 
```{r, echo = FALSE, message = FALSE, warning = FALSE}
juvenile_weight_comp <- juvenile_data %>% 
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm = T),
            sd_weight = sd(weight, na.rm = T),
            n = n())

kbl(juvenile_weight_comp, col.names = c("Sex",
                                        "Mean weight",
                                        "Standard deviation",
                                        "Sample size")) %>% 
  kable_styling(bootstrap_options = c("striped"))
```

```{r, include =FALSE}
# Get vectors for t-test
juveniles_f <- juvenile_data %>% 
  filter(sex == "f") %>% 
  pull(weight)

juveniles_m <- juvenile_data %>% 
  filter(sex == "m") %>% 
  pull(weight)
```

```{r, include = FALSE}
# Means and standard deviation of both groups

male_mean <- mean(juveniles_m)
male_sd <- sd(juveniles_m)

female_mean <- mean(juveniles_f, na.rm = T)
female_sd <- sd(juveniles_f, na.rm = T)
```

```{r, include = FALSE}
# T-test and effect size

juvenile_hares_ttest <- t.test(juveniles_f, juveniles_m, na.rm = T)
juvenile_hares_ttest

juvenile_effsize <- cohen.d(juveniles_m, juveniles_f, na.rm = T)
```

On average, juvenile male snowshoe hares are bigger than juvenile female snowshoe hares (`r round(male_mean,2)` $\pm$ `r round(male_sd, 2)` and `r round(female_mean,2)` $\pm$ `r round(female_sd, 2)` mm, respectively; mean $\pm$ 1 standard deviation). While the absolute difference in means is `r round(male_mean, 2) - round(female_mean, 2)` mm (a `r round((male_mean - female_mean)/((male_mean + female_mean)/2)*100, 2)`% difference), the difference in means is not significant (Welch's two-sample t-test: t(`r round(juvenile_hares_ttest$parameter,2)`) = `r round(juvenile_hares_ttest$statistic,2)`, p > 0.001), and the effect size is small (Cohen's *d* = `r round(juvenile_effsize$estimate,2)`).

### Relationship between juvenile weight and hind foot length

```{r, include = FALSE}
# Exploring linearity
ggplot(data = juvenile_data, aes(x = hindft, y = weight)) +
  geom_point()
```

```{r, include =  FALSE}
# Linear model
juvenile_hares_lm <- lm(weight ~ hindft, data = juvenile_data)

summary(juvenile_hares_lm)
```

```{r, include = FALSE}
# Access model outputs
juvenile_hares_lm_tidy <- tidy(juvenile_hares_lm)

juvenile_hares_int <- juvenile_hares_lm_tidy$estimate[1]
juvenile_hares_int

juvenile_hares_coef <- juvenile_hares_lm_tidy$estimate[2]
juvenile_hares_coef
```

```{r, include = FALSE}
# Access other model info
juvenile_hares_out <- glance(juvenile_hares_lm)
juvenile_hares_out

plot(juvenile_hares_lm)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Visualizing the model
ggplot(data = juvenile_data, aes(x = hindft, y = weight)) +
  geom_point(size = 1.5) +
  geom_smooth(method = "lm",
              color = "red",
              size = 0.5,
              fill = "gray10",
              alpha = 0.5) +
  labs(x = "Hind foot length (mm)", y = "Weight (g)") +
  theme_light() +
  stat_regline_equation(label.x = 60, label.y = 1500)
```

```{r, include = FALSE}
# Pearson's r
juvenile_hares_cor <- cor.test(juvenile_data$hindft, juvenile_data$weight)

juvenile_hares_cor
```

**Figure 3.** Relationship between hind foot length (mm) and weight (g) for juvenile snowshoe hares. Linear model summary: $\beta$~1~ = `r round(juvenile_hares_lm_tidy$estimate[2], 2)` g mm^-1^, p < 0.001, R^2^ = `r round(juvenile_hares_out$r.squared,2)`, Pearson's *r* = `r round(juvenile_hares_cor$estimate,2)`). Data: Kielland et  al. (2017).

The relationship between hind foot length and weight does not appear to be linear for juvenile snowshoe hares (Figure ). Simple linear regression revealed that hind foot length does not predict juvenile snowshoe hare weight (p < 0.001, R^2^ = `r round(juvenile_hares_out$r.squared,2)`) with an average slope of $\beta$ = `r round(juvenile_hares_lm_tidy$estimate[2], 2)` g mm^-1^ (i.e., for each one millimeter increase in hind foot length we expect an average increase in body mass of `r round(juvenile_hares_lm_tidy$estimate[2], 2)` g). Hind foot length and weight are moderately positively correlated (Pearson's *r* = `r round(juvenile_hares_cor$estimate,2)`, p < 0.001). Diagnostic plots (not included) reveal non-normally distributed and heteroscedastic residuals.

I don't think a linear model described the relationship between hind foot length and weight well because of heteroscedasticity and non-linear QQ-plot results.

## Summary

Exploratory data analysis reveals the following initial findings:

Male juvenile snowshoe hares are larger than female juvenile snowshoe hares.
Male juvenile snowshoe hares are larger in terms of weight (g) than female snowshoe hares (the difference is not significant, and the effect size is small)
Hind foot length does not appear to be linearly related with weight across juvenile snowshoe hares. The relationship (by simple linear regression) is significant, with a slope (expected average increase in body mass with a 1 mm increase in flipper length) of 49.69 g mm-1.


## Citation 

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative.
